Partial Least Squares Regression with Multiple Functional and Scalar Predictors

### 2025-07-10 manuscript/main.tex
direction computation new notation

Given an orthonormal basis $\{b_l(t)\}$ of $L^2([0,1])$, the functional predictors and the corresponding regression coefficients can be decomposed into
\begin{equation*}
	X_{ij}(t) = \sum_{l=1}^\infty \theta_{ijl} \, b_l(t), \quad 
	\beta_j(t) = \sum_{l=1}^\infty \eta_{0,jl} \, b_l(t),
\end{equation*}
where $\theta_{ijl}$ and $\eta_{0,jl}$ are the coefficients of $X_{ij}(t)$ and $\beta_j(t)$ corresponding to the $l$th basis function $b_l(t)$, respectively.
In practice, functional predictors are observed only at a finite set of time points rather than as complete curves. Rather than reconstructing infinite-dimensional object stated above via finite observations, we approximate each predictor using a finite-dimensional projection by truncating its basis expansion at a finite number $M$. This $M$ can be set to a sufficiently large value (e.g., 15 or 20) to capture a wide range of functional patterns. 
Notably, it is unnecessary to fine-tune $M$, as a  penalization scheme introduced in Section \ref{sec: Regularization of PLS Components} regularizes each functional component to achieve the desired smoothness. We denote the projected predictors and coefficients as
\begin{equation*}
	\tilde{X}_{ij}(t) = \sum_{l=1}^M \theta_{ijl} \, b_l(t), \quad 
	\tilde{\beta}_j(t) = \sum_{l=1}^M \eta_{0,jl} \, b_l(t).
\end{equation*}
Write $\boldsymbol{\theta}_{ij} := (\theta_{ij1}, \ldots, \theta_{ijM})^\top$ and  $\boldsymbol{\eta}_{0j} := (\eta_{0,j1}, \ldots, \eta_{0,jM})^\top$. 
Let $\Theta_j$ be an $n \times M$ matrix whose rows are given by $\{\boldsymbol{\theta}_{ij}, \; i = 1, \ldots, n\}$.

@@

	\subsubsection{Smoothed estimation of PLS components}\label{sec: pls component computation intermediate}
	Building on the basis function approximation from Section~\ref{sec: finite basis approximation}, we now introduce an intermediate strategy for implementing the first step of the PLS algorithm. This approach directly obtains a hybrid PLS direction, sharing the same structure as the hybrid predictors presented in \eqref{eq: Hybrid functional model}, by simply solving a linear equation.

We now describe the strategy for computing the PLS direction at the $l$-th iteration. For simplicity, we omit the iteration index and denote the residualized response by $Y$, with finite-sample observations given by $\mathbf{y} = (y_1, \ldots, y_n) \in \mathbb{R}^n$. Let $\widetilde{W} \in \mathcal{H}$ denote the residualized, basis-approximated hybrid predictor, and let $\widetilde{\mathbf{C}}^{(1)}, \ldots, \widetilde{\mathbf{C}}^{(k)} \in \mathbb{R}^{n \times M}$ denote the basis coefficient matrices of its functional part in the finite-sample observations. Our strategy seeks the direction $\boldsymbol{\xi} \in \mathcal{H}$, whose functional parts lying in 
$\operatorname{span}(\Phi^{(1)}), \ldots, \operatorname{span}(\Phi^{(K)})$
, that maximizes the empirical covariance between the resulting PLS component $\langle \widetilde{W}, \boldsymbol{\xi} \rangle_{\mathcal{H}}$ and the response.

For the computation of the estimated PLS direction  $\hat{\boldsymbol{\xi}}$, 
we obtain the basis coefficients of its functional part, denoted by $\mathbf{d}^{(1)}, \ldots, \mathbf{d}^{(K)} \in \mathbb{R}^{M}$,
and the scalar part  
$\mathbf{d} \in \mathbb{R}^p$,
by solving linear systems
\begin{equation*}
 \boldsymbol{\Phi}^{(1)}
\mathbf{d}^{(1)} 
=
(\widetilde{\mathbf{C}}^{(1)} \boldsymbol{\Phi}^{(1)})^\top
\mathbf{y},
\ldots
,
\boldsymbol{\Phi}^{(K)}
\mathbf{d}^{(K)} 
=
(\widetilde{\mathbf{C}}^{(K)} \boldsymbol{\Phi}^{(K)})^\top
\mathbf{y},
~\text{and}~
\mathbf{d} := \mathbf{Z}^\top \mathbf{y} .
\end{equation*}
The estimated PLS direction is the tuple $\hat{\boldsymbol{\xi}} \in \mathcal{H}$, defined with these coefficients and normalized with respect to the 
hybrid Hilbert norm (Definition
\ref{def:hilbert_space})
as follows:
\begin{equation}\label{def:computed_PLS_direction}
	\hat{\boldsymbol{\xi}} :=
	\frac{1}{
	\bigl(
	\mathbf{d}^\top \mathbf{d} + 
	\sum_{k=1}^K (\mathbf{d}^{(k)})^\top  \boldsymbol{\Phi}^{(K)}  \mathbf{d}^{(k)} 
	\bigr)^{1/2}
	}
	\biggl(
	\sum_{m=1}^M d^{(1)}_m \phi^{(1)}_m,
	\ldots,
	\sum_{m=1}^M d^{(K)}_m \phi^{(K)}_m,
	\mathbf{d}
	\biggr).
\end{equation}
The details of this hybrid norm computation are provided in Appendix~\ref{section:proof:eigenproblem_derivation}.
Finally, the estimated $l$th PLS scores for the $n$ observations,
denoted $\hat{\boldsymbol{\rho}} \in \mathbb{R}^n$
 is computed via hybrid inner product. In a matrix multiplication form, it is computed as:
	\begin{align*}
		\hat{\boldsymbol{\rho}}
		&:=
\biggl(
	\langle \widetilde{\mathbf{W}}_1, \hat{\boldsymbol{\xi}} \rangle_{\mathcal{H}},
	\ldots,
		\langle \widetilde{\mathbf{W}}_n, \hat{\boldsymbol{\xi}} \rangle_{\mathcal{H}}
\biggr)^\top
\\&	=
	\frac{1}{
	\bigl(
	\mathbf{d}^\top \mathbf{d} + 
	\sum_{k=1}^K (\mathbf{d}^{(k)})^\top  \boldsymbol{\Phi}^{(K)}  \mathbf{d}^{(k)} 
	\bigr)^{1/2}
}
	\biggl(
	\mathbf{Z} \mathbf{d}
	+
	\sum_{k=1}^K \widetilde{\mathbf{C}}^{(k)} \boldsymbol{\Phi}^{(k)} \mathbf{d}^{(k)}
	\biggr)
	. 
\end{align*}
The details of this computation are provided in Appendix~\ref{section:proof:eigenproblem_derivation}.
Although the unnormalized coefficients are computed separately for the functional and scalar part of the PLS direction, the normalization step couples them, allowing the components to influence each other. This enables the procedure to account for the correlation between functional and scalar components.

%	consider the basis function expansions of the functional predictors $X_i^{(k)}(t_k) = \sum_{m=1}^M c_{im}^{(k)} b^{(k)}_{m}(t_k)$, $i=1, \ldots, n$, and those of the functional parts of the PLS component $\psi^{(k)}(t_k) = \sum_{m=1}^M a^{(k)}_{m} b^{(k)}_m(t_k)$, $k=1, \ldots, K$. Let 
\begin{proposition} \label{Proposition: generalized rayleigh quotient}
	Let $\boldsymbol{\xi} \in \mathcal{H}$ be any unit-norm direction such that its $1$st through $K$th functional components lie in $\operatorname{span}(\Phi^{(1)}), \ldots, \operatorname{span}(\Phi^{(K)})$, respectively.
	Then, the empirical covariance between the projected component $\langle \widetilde{\mathbf{W}}, \boldsymbol{\xi} \rangle_{\mathcal{H}}$ and the response $Y$, approximated in terms of the basis, is given by
	\begin{equation*}
		\widehat{\textnormal{Cov}} \left( \langle \widetilde{\mathbf{W}}, \boldsymbol{\xi} \rangle_{\mathcal{H}}, Y \right)
		:=
		\frac{1}{n} 
		\sum_{i=1}^n
		y_i
		\langle \widetilde{\mathbf{W}}_i, \boldsymbol{\xi} \rangle_{\mathcal{H}}.
	\end{equation*}
Then the PLS direction 
$	\hat{\boldsymbol{\xi}}$ 
 computed in \eqref{def:computed_PLS_direction} maximizes $\widehat{\textnormal{Cov}} (\langle \widetilde{\mathbf{W}}, \boldsymbol{\xi} \rangle_{\mathcal{H}}, Y )$.
	\end{proposition}

>>>>

Given an orthonormal basis $\{b_l(t)\}$ of $L^2([0,1])$, the functional predictors and the corresponding regression coefficients can be decomposed into
\begin{equation*}
	X_{ij}(t) = \sum_{l=1}^\infty \theta_{ijl} \, b_l(t), \quad 
	\beta_j(t) = \sum_{l=1}^\infty \eta_{0,jl} \, b_l(t),
\end{equation*}
where $\theta_{ijl}$ and $\eta_{0,jl}$ are the coefficients of $X_{ij}(t)$ and $\beta_j(t)$ corresponding to the $l$th basis function $b_l(t)$, respectively.
In practice, functional predictors are observed only at a finite set of time points rather than as complete curves. Rather than reconstructing infinite-dimensional object stated above via finite observations, we approximate each predictor using a finite-dimensional projection by truncating its basis expansion at a finite number $M$. This $M$ can be set to a sufficiently large value (e.g., 15 or 20) to capture a wide range of functional patterns. 
Notably, it is unnecessary to fine-tune $M$, as a  penalization scheme introduced in Section \ref{sec: Regularization of PLS Components} regularizes each functional component to achieve the desired smoothness. We denote the projected predictors and coefficients as
\begin{equation*}
	\widetilde{X}_{ij}(t) = \sum_{l=1}^M \theta_{ijl} \, b_l(t), \quad 
	\tilde{\beta}_j(t) = \sum_{l=1}^M \eta_{0,jl} \, b_l(t).
\end{equation*}
The $i$th  hybrid predictor with projected functional parts, is a tuple, denoted as
\begin{equation*}
	\widetilde{W}_i := 
	(
	X_{i1}, \ldots, X_{iK}, \boldsymbol{Z}_i
	).
\end{equation*}
We adopt the following notion for the coefficients:
\begin{equation}
	 \boldsymbol{\theta}_{ij} := (\theta_{ij1}, \ldots, \theta_{ijM})^\top \in \mathbb{R}^M,
	 \quad
	 \boldsymbol{\eta}_{0j} := (\eta_{0,j1}, \ldots, \eta_{0,jM})^\top  \in \mathbb{R}^M,
\end{equation}
and
\begin{equation}
	 \Theta_j := (\boldsymbol{\theta}_{1j}, \ldots, \boldsymbol{\theta}_{nj})^\top  \in \mathbb{R}^{n \times M}.
\end{equation}

@@

	\subsubsection{Smoothed estimation of PLS components}\label{sec: pls component computation intermediate}
	Building on the basis function approximation from Section~\ref{sec: finite basis approximation}, we now introduce an intermediate strategy for implementing the first step of the PLS algorithm. This approach directly obtains a hybrid PLS component direction, sharing the same structure as the hybrid predictors presented in \eqref{eq: Hybrid functional model}, by simply solving a linear equation.
We now describe the strategy for computing the PLS component direction at the $l$-th iteration. For simplicity, we omit the iteration index and assume that the observations have been residualized and are orthogonal to those from previous steps (see Section~\ref{subsec: Step 2: Computing Predictors via Hybrid-on-Scalar Regression} for the residualization procedure).

The goal is to find a unit-norm direction $\boldsymbol{\xi} \in \mathcal{H}$, whose functional components lie in $\operatorname{span} \bigl( b_1(t), \ldots, b_M(t) \bigr)$, that maximizes the empirical covariance between the resulting PLS component scores $\langle \widetilde{W}_1, \boldsymbol{\xi} \rangle_{\mathcal{H}}, \ldots, \langle \widetilde{W}_n, \boldsymbol{\xi} \rangle_{\mathcal{H}}$ and the responses $y_1, \ldots, y_n$.
We denote the solution as
\begin{equation}
	\hat{\boldsymbol{\xi}} 
	= 
	(\hat{\xi}_1, \ldots, \hat{\xi}_K, \hat{\boldsymbol{\zeta}}),
\end{equation}
where $\hat{\boldsymbol{\zeta}} \in \mathbb{R}^p$ is the scalar part, and each functional part $\hat{\xi}_j \in L^2([0,1])$ lies in $\operatorname{span}(b_1(t), \ldots, b_M(t))$ and is expressed as
\begin{equation*}
	\hat{\xi}_j(t) = 
	\sum_{l=1}^M d_{jl} \, b_l(t), \quad j = 1, \ldots, K.
\end{equation*}

Let $\mathbf{d}_j := (d_{j1}, \ldots, d_{jM})^\top \in \mathbb{R}^M$ and $\mathbf{y} := (y_1, \ldots, y_n)^\top$. We compute $\mathbf{d}_1, \ldots, \mathbf{d}_K$ and $\hat{\boldsymbol{\zeta}}$ as follows.
First, we obtain the unnormalized coefficient vectors. For each $j$, the unnormalized vector $\bar{\mathbf{d}}_j$ is computed by solving
\begin{equation*}
	B \bar{\mathbf{d}}_j = (\Theta_j B)^\top \mathbf{y},
\end{equation*}
 The unnormalized scalar component is given by
\begin{equation}
	\bar{\boldsymbol{\zeta}} := Z^\top \mathbf{y}.
\end{equation}
Next, we normalize all components with respect to the hybrid Hilbert norm (see Definition~\ref{def:hilbert_space}) to ensure $\hat{\boldsymbol{\xi}}$ has unit length. Specifically, we define
\begin{equation}\label{def:computed_PLS_direction}
	\mathbf{d}_j := 
	\frac{1}{
		\left(
		\sum_{j=1}^K \bar{\mathbf{d}}_j^\top B \bar{\mathbf{d}}_j +
		\bar{\boldsymbol{\zeta}}^\top \bar{\boldsymbol{\zeta}}
		\right)^{1/2}
	} \bar{\mathbf{d}}_j, \quad j = 1, \ldots, K,
\end{equation}
and
\begin{equation}
	\hat{\boldsymbol{\zeta}} :=
	\frac{1}{
		\left(
		\sum_{j=1}^K \bar{\mathbf{d}}_j^\top B \bar{\mathbf{d}}_j +
		\bar{\boldsymbol{\zeta}}^\top \bar{\boldsymbol{\zeta}}
		\right)^{1/2}
	} \bar{\boldsymbol{\zeta}}.
\end{equation}

Details of the hybrid norm computation are provided in Appendix~\ref{section:proof:eigenproblem_derivation}.

Finally, the estimated $l$th PLS scores for the $n$ observations,
denoted $\hat{\boldsymbol{\rho}} \in \mathbb{R}^n$
 is computed via hybrid inner product. In a matrix multiplication form, it is computed as:
	\begin{align*}
		\hat{\boldsymbol{\rho}}
		&:=
\biggl(
	\langle \widetilde{\mathbf{W}}_1, \hat{\boldsymbol{\xi}} \rangle_{\mathcal{H}},
	\ldots,
		\langle \widetilde{\mathbf{W}}_n, \hat{\boldsymbol{\xi}} \rangle_{\mathcal{H}}
\biggr)^\top
\\&	=
	\frac{1}{
	\bigl(
	\mathbf{d}^\top \mathbf{d} + 
	\sum_{k=1}^K (\mathbf{d}^{(k)})^\top  \boldsymbol{\Phi}^{(K)}  \mathbf{d}^{(k)} 
	\bigr)^{1/2}
}
	\biggl(
	\mathbf{Z} \mathbf{d}
	+
	\sum_{k=1}^K \widetilde{\mathbf{C}}^{(k)} \boldsymbol{\Phi}^{(k)} \mathbf{d}^{(k)}
	\biggr)
	. 
\end{align*}
The details of this computation are provided in Appendix~\ref{section:proof:eigenproblem_derivation}.
Although the unnormalized coefficients are computed separately for the functional and scalar part of the PLS component direction, the normalization step couples them, allowing the components to influence each other. This enables the procedure to account for the correlation between functional and scalar components.

%	consider the basis function expansions of the functional predictors $X_i^{(k)}(t_k) = \sum_{m=1}^M c_{im}^{(k)} b^{(k)}_{m}(t_k)$, $i=1, \ldots, n$, and those of the functional parts of the PLS component $\psi^{(k)}(t_k) = \sum_{m=1}^M a^{(k)}_{m} b^{(k)}_m(t_k)$, $k=1, \ldots, K$. Let 
\begin{proposition} \label{Proposition: generalized rayleigh quotient}
	Let $\boldsymbol{\xi} \in \mathcal{H}$ be any unit-norm direction such that its $1$st through $K$th functional components lie in $\operatorname{span} \bigl( b_1(t), \ldots, b_M(t) \bigr)$, respectively.
	Then, the empirical covariance between the projected component $\langle \widetilde{\mathbf{W}}, \boldsymbol{\xi} \rangle_{\mathcal{H}}$ and the response $Y$, approximated in terms of the basis, is given by
	\begin{equation*}
		\widehat{\textnormal{Cov}} \left( \langle \widetilde{\mathbf{W}}, \boldsymbol{\xi} \rangle_{\mathcal{H}}, Y \right)
		:=
		\frac{1}{n} 
		\sum_{i=1}^n
		y_i
		\langle \widetilde{\mathbf{W}}_i, \boldsymbol{\xi} \rangle_{\mathcal{H}}.
	\end{equation*}
Then the PLS component direction 
$	\hat{\boldsymbol{\xi}}$ 
 computed in \eqref{def:computed_PLS_direction} maximizes $\widehat{\textnormal{Cov}} (\langle \widetilde{\mathbf{W}}, \boldsymbol{\xi} \rangle_{\mathcal{H}}, Y )$.
	\end{proposition}
